Space astrometry of the very massive $\sim$$150\,M_\odot$ candidate runaway star VFTS682

### 2018-08-22 VFTS682_APJL.tex
MNRAS Letter ready for submission

How very massive stars form is still an intriguing open question in
stellar astrophysics.  VFTS682 is among the most massive stars known,
with an inferred initial mass of $\sim$$150\,M_\odot$. It is located
in 30 Doradus at a projected distance of 29\,pc from the central
cluster R136. % Its apparent isolation
The absence of other massive stars in its immediate vicinity
led to two intriguing hypotheses: % posed in earlier work:
either it formed in relative isolation % through a new mode of star formation
or it was ejected dynamically from the central cluster.

% Aiming to shine light on this debate,
In this context, 
we investigate the kinematics of
VFTS682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space
  Telescope} astrometry. We derive a projected velocity relative to
the cluster of $38\pm17\kms$ ($1\sigma$ confidence interval).  The direction and magnitude of
the proper motion are consistent with VFTS682 being a slow runaway
ejected from the central cluster, given its distance and the inferred
age. The \todo{rephrase positive, see Fabian and Roeland's comments}

However, the error bars are substantial and the proper motion measurements alone cannot rule out the counter hypothesis with high confidence.

If future data confirm the runaway nature, this would make VFTS682
the most massive runaway star known to date. This hypothesis is
further supported by a variety of
circumstantial clues. The central cluster is known to harbor several
other stars more massive than $150\,M_\odot$ of similar in spectral type.
The very massive O stars VFTS16 (and possibly VFTS72), which has recently been proposed as a dynamically ejected runaway, provides direct evidence that R136 is indeed capable of ejecting some of its most massive members, consistent with the predictions of numerical simulations.
\end{abstract}

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2D}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta_{Gaia}$  &[degrees] &  $14_{-31}^{+36}$  & (1,5)\\  
      % &$v_\mathrm{2D, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2D, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta_\mathrm{HST}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                  %        &$v_\mathrm{2D, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
    \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from
      R136 at age zero}} \\
      \hline
      &$v_\mathrm{2D}$  &[$\kms$] & $29\pm 6$ & (2) \\  
      &$\theta$  &[degrees] &  $\sim0$  & \\ 
       \hline

@@

The V-band light curve of VFTS682  shows
variations at a $\sim$10\% level on a timescale of years, which is
unusual for Wolf-Rayet stars and more typical for Luminous Blue
Variable (LBV) stars \citep{udalski:08, bestenlehner:11}. It also
shows a mid-infrared excess \citep{gruendl:09}.

@@

The black points in~\Figref{fig:dist} show the proper motion 
(left panel) and the projected flight direction (right panel) relative
to R136 of
VFTS682. For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS O/B-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. The green vertical bands highlight the expectations for these
two quantities if VFTS682 were ejected from R136 early in its
lifetime. Although the error bars are substantial and VFTS682 is not
an outlier compared to other OB-type and Wolf-Rayet stars, the agreement
suggests that the star is indeed a ``slow-runaway'' as suggested by
\cite{bestenlehner:11}.

% Figure~\ref{fig:dist} illustrates the kinematic properties of VFTS OB-type
% and Wolf-Rayet stars in the 30 Doradus region with \emph{Gaia} DR2 errors on the proper motion
% components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
% including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
% blue). The green vertical bands show the expectations for VFTS682 if
% it were ejected from R136 early in its lifetime. We highlight this star together with the two O-type
% runaways, VFTS16 and VFTS72, \mbox{(re-)identified} by \cite{lennon:18}.

% The left panel shows the distribution in proper motion relative
% to R136 from \emph{Gaia} DR2, selecting stars with small error bars only.
% Although VFTS682 is not peculiar, its relative proper motion
% falls in the range expected for it to reach its present day isolation
% within its apparent age (green vertical band). This suggests it is indeed a ``slow-runaway'' as
% suggested by \cite{bestenlehner:11}.
Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other, but also with no relative motion
within $2\sigma$.

% The left panel shows the distribution in two-dimensional proper motion relative
% to R136, and shows that, although VFTS682 is not a peculiar star in terms of
% its motion relative to R136. Subtracting the mean motion of the field,
% we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
% $\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
% are consistent with each other and indicate
% a non-zero peculiar projected motion that would make it a slow runaway
% star as expected by \cite{bestenlehner:11} if the star was ejected
% from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
% More importantly, both the proper motion
% measurement from \emph{Gaia} and HST fall in the range expected to
% reach its present day location % at $\sim$29\,pc from the cluster core
% if the
% star was ejected very early in its evolution. This range is
% highlighted by the green vertical band, and its width is determined by
% the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

% The right panel shows the distributions in projected flight direction for these
% stars: $\theta$ is the angle between the projected radial direction from the
% center of R136 to each star and its projected relative proper motion (see also \Figref{fig:main}). Dynamical ejections from the
% cluster should %are expected to
% produce
% close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
% % the range of angles corresponding to
% an opening angle around the
% projected radial direction of 45\,degrees. The direction of motion of VFTS682 % is
% % poorly constrained because of the large errors on the proper motion
% % components both in HST and \emph{Gaia} data, but
% % it
% is consistent with dynamical ejection from R136.
We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. We also show VFTS16 and VFTS72 for completeness
\citep[see][]{lennon:18}. The thick yellow arrows are proportional to
the relative proper motion from \emph{Gaia} DR2, and the thick green
line illustrates the relative proper motion from HST. The error cone
on the direction of motion is illustrated by the corresponding
prolongation in the direction opposite to the motion, and we also show
the most likely origin of the stars accounting for their apparent age.
This figure illustrates that R136 is the most likely origin of these stars, although the large error bars
prevent a robust identification for VFTS682, and there is some tension
between the apparent age and the present day distance from the cluster
core for VFTS16 and VFTS72 \citep[][]{lennon:18}.

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[][]{bestenlehner:11}. % \citep[corresponding to $\sim$29\,pc at a
% distance of 
% 50\,kpc,][]{bestenlehner:11}.
% As in the rest of this study, we neglect for
% simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is consistent with a very early
ejection from the cluster (see \Tabref{tab:star_param}).
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

Based on our results, we consider likely that VFTS682 is the most massive
(slow) runaway known to date, with a two-dimensional
projected velocity with respect to R136 of
$\sim38\kms$ (averaging the \emph{Gaia} DR2 and HST
results), but compatible with zero within $2\sigma$. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.

If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. If the
cluster age \citep[$\lesssim2$\,Myr,][]{crowther:10, sabbi:12} is
indeed smaller than the shortest stellar lifetime
\citep[$\sim$3\,Myr,][]{brott:11,kohler:15, zapartas:17}, the ejection of VFTS682
from the disruption of a binary is excluded. %  since the region is too young for stars
% to have experienced core-collapse already. 
The kinematic age we infer is smaller than the kinematic age of
$\sim$1.5\,Myr for VFTS16 found in \cite{lennon:18}, which indicates
that VFTS682 was ejected later than VFTS16, potentially at a time similar to VFTS72.

@@

The simulations of \cite{fujii:11} suggest that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other
stars. The spectral similarities between VFTS682 and stars in the core
of R136 are in agreement with this ``bully binary'' model. Interpreting the kinematics of VFTS682 through the lens of their simulations
suggests the presence of a close binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. The kinematic age of VFTS682 puts an
upper limit to the timescale to form the ``bully binary'' in
R136. Such a binary could be R145 according to \cite{fujii:11}, and it might be an
ideal observational candidate for a dynamically formed progenitor system of
a binary black-hole, provided that stars this massive can avoid a
pair-instability supernova \cite[e.g.,][]{rakavy:67} at LMC
metallicity \citep[see also][]{langer:07}.

>>>>

How very massive stars form is still an intriguing open question in
stellar astrophysics.  VFTS682 is among the most massive stars known,
with an inferred initial mass of $\sim$$150\,M_\odot$. It is located
in 30 Doradus at a projected distance of 29\,pc from the central
cluster R136. % Its apparent isolation
The absence of other massive stars in its immediate vicinity
led to two intriguing hypotheses: % posed in earlier work:
either it formed in relative isolation % through a new mode of star formation
or it was ejected dynamically from the central cluster. 
In this context, 
we investigate the kinematics of
VFTS682 as obtained by \emph{Gaia} and multi-epoch \emph{Hubble Space
  Telescope} astrometry. We derive a projected velocity relative to
the cluster of $38\pm17\kms$ ($1\sigma$ confidence interval). Although
the error bars are substantial, two completely independent measures
suggest that VFTS682 might be a slow runaway
ejected from the central cluster, given its distance and the inferred
age. This hypothesis is further supported by a variety of
circumstantial clues. The central cluster is known to harbor several
other stars more massive than $150\,M_\odot$ of similar in spectral
type. The very massive O stars VFTS16 (and possibly VFTS72), which has
recently been proposed as a dynamically ejected runaway, provides
direct evidence that R136 is indeed capable of ejecting some of its
most massive members, consistent with the predictions of numerical
simulations. If future data confirm the runaway nature, this would make VFTS682
the most massive runaway star known to date. 
\end{abstract}

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.10\pm0.08$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2D}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta_{Gaia}$  &[degrees] &  $14_{-31}^{+36}$  & (1,5)\\  
      % &$v_\mathrm{2D, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2D, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta_\mathrm{HST}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                  %        &$v_\mathrm{2D, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
      \hline
      \multicolumn{5}{l}{\emph{Weighted average relative proper motion
      for VFTS682}}\\
      \hline
      &$\delta\mu_\mathrm{RA, avg}$  &[$\masyr$] & $0.08\pm0.07$ & (5) \\
      &$\delta\mu_\mathrm{DEC, avg}$  &[$\masyr$] & $0.14\pm0.07$ &  (5) \\
       &$\delta\mu_\mathrm{avg}$  &[$\masyr$] & $0.16\pm0.07$ &  (5) \\
      &$v_\mathrm{2D, avg}$  &[$\kms$] & $38\pm17$ & (5)\\
      % &$\theta_\mathrm{avg}$  &[degrees] &  \todo{TBD} $_{-}^{+}$   & (5)\\  
      \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from
      R136 at age zero}} \\
      \hline
      &$v_\mathrm{2D}$  &[$\kms$] & $29\pm 6$ & (2) \\  
      &$\theta$  &[degrees] &  $\sim0$  & \\ 
       \hline

@@

The V-band light curve of VFTS682  shows
variations at a $\sim$10\% level on a timescale of years, which is
unusual for Wolf-Rayet stars and more typical for Luminous Blue
Variable (LBV) stars \citep{parker:93, udalski:08, bestenlehner:11}. It also
shows a mid-infrared excess \citep{gruendl:09}.

@@

The black points in~\Figref{fig:dist} show the proper motion 
(left panel) and the projected flight direction (right panel) relative
to R136 of
VFTS682 from \emph{Gaia} DR2. $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its projected relative proper motion
(see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce close to radial ejections, i.e.~$\theta\simeq0$. The green
vertical bands highlight the expectations for these two quantities if VFTS682 were ejected from R136 early in its
lifetime. For comparison, we also show the relative
proper motion of VFTS16 and VFTS72 (gray points), and the distribution
in relative proper motion and flight direction for all the VFTS OB-type and Wolf-Rayet stars with
\emph{Gaia} DR2 errors on the proper motion
components of less than $0.1\,\mathrm{mas\ yr^{-1}}$ (dark blue lines,
including VFTS682), and less than $0.05\,\mathrm{mas\ yr^{-1}}$ (light hatched
blue). Although the error bars are substantial and VFTS682 is not
an outlier compared to other OB-type and Wolf-Rayet stars, the agreement
suggests that the star is indeed a ``slow-runaway'' as suggested by
\cite{bestenlehner:11}.

Subtracting
the mean motion of R136, we obtain relative proper motions (projected
velocities) of $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both values
are consistent with each other, but also with no relative motion
within $2\sigma$. The weighted average of these two independent
measurements is $\delta \mu_\mathrm{avg}=0.16\pm0.07\masyr
(38\pm17\kms)$.

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. We also show VFTS16 and VFTS72 for completeness
\citep[see][]{lennon:18}. The thick yellow arrows are proportional to
the relative proper motion from \emph{Gaia} DR2, and the thick green
line illustrates the relative proper motion from HST. The error cone
on the direction of motion is illustrated by the corresponding
prolongation in the direction opposite to the motion, and we also show
the most likely origin of the stars accounting for their apparent age.
This figure illustrates that R136 is the most likely origin of these stars, although the large error bars
prevent a robust identification for VFTS682, and there is some tension
between the apparent age and the present day distance from the cluster
core for VFTS16 and VFTS72 \citep[][]{lennon:18}. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VFTS682) has
large peculiar radial velocity and therefore accurate distances along
the line of sight are
needed to contrain the flight direction in three dimensions.

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_\mathrm{avg}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.16\masyr} \simeq 0.7\pm\,0.3\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[][]{bestenlehner:11}. % \citep[corresponding to $\sim$29\,pc at a
% distance of 
% 50\,kpc,][]{bestenlehner:11}.
% As in the rest of this study, we neglect for
% simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is consistent with a very early
ejection from the cluster (see \Tabref{tab:star_param}).
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

Based on our results, we consider likely that VFTS682 is the most massive
(slow) runaway known to date, with a two-dimensional
projected velocity with respect to R136 of
$38\pm17\kms$ (averaging the \emph{Gaia} DR2 and HST
results). Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.

If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.7\pm0.3$\,Myr ago. If the
cluster age \citep[$\lesssim2$\,Myr,][]{crowther:10, sabbi:12} is
indeed smaller than the shortest stellar lifetime
\citep[$\sim$3\,Myr,][]{brott:11,kohler:15, zapartas:17}, the ejection of VFTS682
from the disruption of a binary is excluded. %  since the region is too young for stars
% to have experienced core-collapse already. 
The kinematic age we infer is smaller than the kinematic age of
$\sim$1.5\,Myr for VFTS16 found in \cite{lennon:18}, which indicates
that VFTS682 was ejected later than VFTS16, and potentially later than
VFTS72 too.

@@

The simulations of \cite{fujii:11} suggest that
early in the evolution of a cluster, dynamical interactions form an extremely
massive binary, which then tightens its orbit by ejecting other
stars. The spectral similarities between VFTS682 and stars in the core
of R136 are in agreement with this ``bully binary'' model. Interpreting the kinematics of VFTS682 through the lens of their simulations
suggests the presence of a close binary with total mass
$M_1+M_2\gtrsim 300\,M_\odot$ in the core of the cluster. The kinematic age of VFTS682 puts an
upper limit to the timescale to form the ``bully binary'' in
R136. If identified, such a binary % could be R145 according to \cite{fujii:11}, and it
might be an ideal observational candidate for a dynamically formed progenitor system of
a binary black-hole, provided that stars this massive can avoid a
pair-instability supernova \cite[e.g.,][]{rakavy:67} at LMC
metallicity \citep[see also][]{langer:07}.
